Tidal streams in triaxial systems

### 2015-02-14 ms.tex
more intro work

On the largest scales ($\gtrsim$30~kpc for a Milky-Way-like galaxy) in the halos of galaxies, the gravitational potentials are thought to be triaxial. However, despite suggestive evidence, this fundamental prediction from LCDM has not been conclusively confirmed from observations. Around other galaxies, it is generally hard to measure the 3D shape of [..] because they are seen in projection. From our position within the Milky Way, our 3D view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. Unfortunately, the halo also has a low density of tracers. Luckily, however, many of the halo stars are likely associated with debris stripped from in-falling satellite galaxies.

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. The phase-space density and therefore the observable properties of the tidal debris are sensitive to the internal progenitor properties, orbit of the progenitor, and underlying mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions) of the debris and the orbit of the progenitor system, both of which are determined by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances. By simultaneously modeling the observed density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these large scales.

Tidal streams are generally modeled as 1D structures with simple internal progenitor dynamics and density evolution. 
Pearson et al. (2014) tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in an oblate and a triaxial potential using \emph{Streakline} models (K\"upper et al. (XXXX)) : 
		\item in the oblate potential, a thin model stream is easily found that reproduces the observed stellar density morphology of the stream;
		\item in the triaxial potential, the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream, given the observational constraints of the present-day position and velocity of the cluster.
This result suggests that the shape of the host galaxy potential directly influences the morphology of the resulting debris, and that other morphological classes exist between thin, 1D streams and diffuse, 2D shells.
To understand the formation of these fans and the implications of this work we must first understand the complex orbital structure of triaxial systems.

In this work, we study the effect of orbital classes that are prevalent in triaxial systems (e.g., stochastic, near-resonant, resonant, and quasiperiodic) on the resulting morphology of tidal streams.

For (typically) many orbits, these tidally stripped stars remain spatially coherent.
Of particular interest are cold, young dynamical structures --- ``tidal streams'' -- which provide useful information about the mass distributions at large distances in the galaxies within which they orbit. [At these distances, disk / classic baryonic components can't help you, mostly dark matter]
The formation and morphology of tidal streams has been studied extensively in a variety of assumed potential forms. [Stream formation in oblate, spherical potentials, action-angle, etc.]

Adding third degree of freedom opens up many new orbits / resonances / etc. -- Valluri and Merritt work...

The halos of galaxies are filled with substructure...

For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus.

>>>>

On the largest scales ($\gtrsim$30~kpc for a Milky-Way-like galaxy) in the halos of galaxies, gravitational potentials are thought to be triaxial. However, despite suggestive evidence, this fundamental prediction from $\Lambda$CDM cosmology has not been conclusively confirmed from observations. Around other galaxies, it is generally hard to measure the 3D mass profile because they are seen in projection. From our position within the Milky Way, our view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. Unfortunately, the Milky Way halo has a low density of visible tracers. Luckily, however, many of the halo stars are likely associated with debris stripped from in-falling satellite galaxies.

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and underlying mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions) of the debris and the orbit of the progenitor system, both of which are determined by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances ($\sim$10-100~kpc). By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these large scales.

Tidal streams are generally modeled as inherently 1D structures formed on regular orbits with simple internal progenitor dynamics and density evolution. Methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to direct $N$-body simulations. All methods have been tested in some way on simulated observations of data and authors have mainly demonstrated their ability to recover parameters of analytic, static potential forms. Many of the methods have been [...] in axisymmetric or otherwise simple potentials but claim to extend to more complex, generalized potentials (e.g., triaxial of multi-component potentials).

\citet{pearson15} tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} models \citep{kupper12}. They use the observed SDSS number density of stars and a limited number of radial velocities measured for stream members to fit their models to the data. In the oblate potential (a three-component, bulge, disk, and spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential fit to the Sagitarrius stream by \cite{law10}), the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream, given the observational constraints of the present-day position and velocity of the cluster. This result suggests that the shape of the host galaxy potential directly influences the morphology of the resulting debris. In order to extending tidal stream models to more generic potentials, it is important to understand why this stream-fanning occurs and whether this and other strange debris morphologies are generic to more complex potentials.

One obvious difference between the two potentials considered in \citet{pearson15} is the extra symmetry of the oblate potential. For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus.
